Gsmodutils: A python based framework for test-driven genome scale metabolic model development

### 2018-09-27 article/article.tex
Article updates

\title{Gsmodutils: A python based framework for test driven genome scale metabolic model development}

@@

\begin{abstract}
\textbf{Motivation:} Genome scale metabolic models (GSMMs) are increasingly important for systems biology and metabolic engineering research as they are capable of simulating complex steady state phenotypic behaviour.
Constraints based models of this form can include thousands of reactions and metabolites, with many crucial pathways that only become activated in specific simulation settings.
However, despite their widespread use, power and the availability of tools to aid with the construction and analysis of large scale models little methodology is suggested for the management of continually curated large scale models.
This is quickly becoming an issue for industrial systems and synthetic biotechnology applications, which require good quality reusable models integral to the design, build and test cycle.
\\
\textbf{Results:} 
As part of an ongoing effort to improve genome scale metabolic analysis, we have developed a test driven development methodology for the continuous integration of validation data from different sources.
Contributing to the open source technology based around COBRApy, we have developed the \textit{gsmodutils} modelling framework placing an emphasis on test driven design of models through defined test cases.
Crucially, different substrate conditions are configurable allowing users to examine how different designs or curation impact a wide range of system behaviours, minimising error between model versions.\\
\textbf{Availability:} The software framework described within this paper is open source and freely available from \url{http://github.com/SBRCNottingham/gsmodutils} \\
\textbf{Contact:}{james.gilbert@nottingham.ac.uk}
\end{abstract}

\section{Introduction}
Stoichiometric constraints based modelling for biological systems has been a mainstay of systems biology for several decades \cite{fell1986fat, varma1994stoichiometric}.
Given its flexibility, low barrier to entry and requirement only on minimal knowledge regarding the stoichiometry of metabolic networks this structural approach has become an extremely popular method for modelling steady state behaviour of large, biochemical networks \cite{kauffman2003advances}.
Such large scale reconstructions are often referred to as genome scale metabolic models (GSMMs), as the processes is significantly aided through the advent of relatively inexpensive genome sequencing \cite{o2015using, land2015insights}.
Indeed, owing to their ability to model complex aspects of metabolism, GSMMs have been widely adopted as a standard to elucidate and optimise of industrial biotechnology processes \cite{kim2017current}.

@@

As a consequence, a significant amount of work has gone into the management of genome scale models.
The BiGG models database \cite{king2015bigg}, for example, exists to provide a standardised repository of validated models that can be shared and reused.
However, little focus is placed upon the collaborative design aspect of such models with few mechanisms existing to store the \textit{model delta} based tools provided in our software.
Similarly, the MetaNetX \cite{moretti2016metanetx} system exists to provide a standardised namespace and toolchain for GSM analysis.
However, such tools often make it difficult to understand the design decisions made by the initial model authors.

Furthermore, as with many areas of bioinformatic study the number available computational tools has become vast.
This covers huge variety of software platforms including the COBRA toolbox in Matlab \cite{schellenberger2011quantitative}, ScrumPy and COBRApy in Python \cite{poolman2006scrumpy, ebrahim2013cobrapy} with additional tools and libraries such as cameo \cite{cardoso2017cameo}, OptFlux in Java \cite{rocha2010optflux} and SurreyFBA \cite{gevorgyan2010surreyfba}.
Whilst most of these tools are Open Source and follow standards, such as SBML \cite{finney2003systems}, there are often few options that make analysis easy to replicate.
The archetype design, build and test cycle of synthetic biology heavily relies on the usage of bioinformatics software and modelling to improve the production of natural products \cite{carbonell2016bioinformatics}.
In order to speed up the use of bioinformatics tools to produce high value platform chemicals, genome scale models are often used to discover methods for process optimisation \textit{in silico}.

For example, many tools such as RetroPath \cite{carbonell2013retropath}, XTMS \cite{carbonell2014xtms} and GEM-Path \cite{campodonico2014generation} suggest thousands of potential heterologous pathways as well as providing predictions for flux distributions.
Many of these tools significantly increase the value of genome scale models, for example by coupling commodity production to an organism's growth \cite{feist2010model}.
These tools all suggest major changes to wild-type strains must be tracked and compared to allow models to remain relevant.
In effect, mechanisms are required to relate modified models to modified test and production strains.

Similarly, many conventional applications of genome scale models in systems biology have often suffered from unnecessary replication of work due to a lack of standards.
For example, there are now many independently developed models of \textit{Clostridium acetobutylicum} \cite{lee2008genome, senger2008genome, mcanulty2012genome, dash2014capturing, yoo2015quantitative}, an organism used in the production of solvents for over a century.
These models all exist to solve similar biological problems some being updates to the inital base models. 
However, there has been disagreement over fundamental biochemical properties of these anaerobic organisms, notably with the focus on redox balancing \cite{dash2014capturing}.
Indeed, such models also include updates based on improved genome annotations and the inclusion of fluxomic, transcriptomic and metabolomic characterisations \cite{yoo2015quantitative}.

Many of the results presented in original sources are difficult to compare due to a lack of standardisation of reference identifiers for reaction names.
Where models are shared they are often in non-standard spreadsheet formats, rather than SBML models.
Indeed, even in the case of valid SBML models being made available at the time of publication few details are given as to how to run such models.

% Perhaps more worryingly, from an industrial perspective, is that the use of models in bioproduction has not escaped regulatory notice \cite{}.
Furthermore, as models are often used to explain issues with online production and facilitate the optimisation of bioprocessing, it stands to reason that model correctness should attract oversight.
Consequently, methods are needed to ensure that the models themselves are tightly version controlled.
In this paper, we argue for the use of a form of \textit{test driven} model of development, a conceptual approach borrowed from software engineering.
Given the availability of widespread open source virtualisation tools that allow the standardisation of platforms \cite{merkel2014docker}, modellers have the opportunity to share highly reproducible research \cite{chamberlain2014using}.

In this paper we present a software framework geared toward \textit{test driven} genome scale model development, a concept that is taken directly from good software development practices.
By this we mean the notion that, as a model is curated to represent biological phenomena, much of the validation can be turned into specific test cases that can be repeated between model versions.
We provide an example test case for \textit{Clostridium autoethanogenum}, an organism that has had considerable focus in terms of genome scale models and how a working methodology using the software presented here can reduce repetition of work and improve the reproducibility of results.
This article aims to summarise the main objectives of the gsmodutils software and we refer the reader to the software user guide for a more detailed exploration features.

@@

In contrast, an \textit{agile} methodology for the development of models places the focus on adapting work to new requirements \cite{martin2002agile}.
Such an approach best fits genome scale models because they are rarely created to investigate individual processes and, instead, capture the complexity of large systems.
Constraints based models are intrinsically related to available genome annotations.
Such annotations rely heavily on automated matches to related species, with the characterisations of individual genes often being left to a few of critical interest \cite{seemann2014prokka}.
This modelling formation costs an in depth understanding of dynamic behaviour. 
However, capturing steady state phenomena still provides a good understanding of system properties \cite{o2015using}.

@@

Therefore, models undergo significant manual annotation - a process which naturally leads to errors.
In this work we advocate a test driven approach to model development highlighted in Figure \ref{fig:test_driven}.
Here, the model is changed to achieve core research goals that are dynamic in response to the changes of a project.
In order to meet this objective, validation criteria for a model such as growth conditions or the impact of gene knock outs, should be formally set.
When a model is changed, all such validation criteria should be retested to ensure that models do not regress to previous states.

@@

We feel that many of the current software tools for genome scale models do an excellent job of facilitating answers to crucial research and design questions.
However, there is a major gap in terms of the reliability and re-usability of models due to a lack of standardisation and software tools to aid such processes.
The following sections provide an overview of the gsmodutils software framework.
Gsmodutils aims to provide a basis for test driven, version controlled agile model development.
All software is open source and is designed to be interoperable with platforms widely used in the domain of constraints based modelling.

@@

\subsection{Outline and features}
Test driven development is driven by the idea of clearly defined test cases written before significant changes are made to any underlying architecture.
In the case of genome scale models, errors occur as a product of human curation to better represent newly discovered aspects of metabolism.

By automatically integrating COBRApy \cite{ebrahim2013cobrapy} users can easily write convenient test cases following examples given in the user guide.
A standard test case, ensuring that a given model grows on media is given in Figure \ref{fig:test_case}.
When a new model repository is created with the gsmodutils tool, a number of pre-written test cases are automatically added to a file.
However, we stress that the vast majority of individual use cases for a model must be specific to a given biological problem.

@@

\lstset{language=Python}   
\begin{lstlisting}
def test_func(model, project, log):
    status = model.solver.optimize()
    log.assertion(status == 'optimal',
                  ``Works'', # Success message
		  ``Does not work'', # Failure message
                  ``Test description'')
\end{lstlisting}
\label{fig:test_case} 
\caption{An example gsmodutils test case written in python}
\end{figure}

\subsection{Strain designs}
A core aspect behind the implementation of gsmodutils is the concept of a \textit{design}, this encompasses a simple set of changes to a ``wild-type'' model that are required for analysis.
For example a mutant strain that has undergone several gene deletions is not sufficiently different from the original wild-type to merit having its own genome scale model.
However, it is often the case that such deletions are of scientific or industrial interest and, as such, the strain will be used in future work.
Consequently, such designs are hereditary in nature.
By taking the \textit{model delta} between the constraints applied to an initial model and subsequent modifications, gsmodutils allows users to easily reuse and export models with this design.
As designs inherit from a base model, future curation to a wild type model will automatically be included in the designs.
Similarly, designs are self contained and will not interfere with one another allowing a form of project management.
Figure \ref{fig:strain_hered} shows how this could work in a practical situation.

Designs of this nature can also be programmatic, allowing the implementation of features such as non-standard constraints that can be dynamically loaded.
An example of this is shown in Figure \ref{fig:programmatic_design}.
This example converts an existing model to one based on an MILP, as the reaction names do not need to be specified, should reactions be added or removed the process will still apply.
Alternative examples could include reductions of models through methods such as elementary flux modes or minimum cut sets, which can change dramatically with only small changes to stoichiometry.
Furthermore, integrity of all strain designs is automatically tested as part of the default gsmodutils testing framework.

\begin{figure}[ht]
\lstset{language=Python}   
\begin{lstlisting}
def gsmdesign_example_milp(model, project):
    for reaction in model.reactions:
	reaction.variable_kind = "integer"
    return model

\end{lstlisting}
\label{fig:programmatic_design} 
\caption{An example gsmodutils programmatic design written in python. 
This design converts reactions to integer type, allowing an MILP formation.}
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{inheritence2.eps}
  \label{fig:strain_hered}
  \caption{Examples of gsmodutils design inheritance.
  Each design stores the delta between the wild-type base model, any parents and the changes to constraints the design contains.
  In the example presented above a heterologous production pathway is combined with a reusable set of knock-outs.
  In practice, ideally, these designs should relate to real constructs and strains evaluated by experiment.
  }
\end{figure}

\subsection{Development workflow}
In this section we propose a method for the development of genome scale models that integrates gsmodutils with version control systems.
Figure \ref{fig:test_case} highlights the notion of test cases, taken from test driven development.
The outlined workflow is that the user writes a formal test case for some modelling goal, perhaps driven by captured experimental data, that fits a specific form of validation criteria.
We note that, in principle, this test case should be written before changes to a model.

\section{Case study usage Clostridium autoethanogenum}
\textit{Clostridium autoethanogenum} is a bacterial species used for the production of commodity chemicals at industrial scale \cite{abrini1994clostridium,norman2018progress,LT}. A new GSMM of \textit{C. autoethanogenum}, `MetaCLAU', has been analysed to improve this bioprocess (Norman et al., submitted for peer review).
In this section we describe how gsmodutils has been utilised to ensure that future versions of MetaCLAU will remain functionally relevant from the perspective of industrial biotechnology.

@@

\textbf{Minimisation of flux based tests:}
One conventional approach in FBA is to set an optimisation criterion of minimising flux across enzymatic reactions with a fixed biomass constraint \cite{holzhutter2006thegeneralised}. 
The solution to this FBA problem represents minimal protein investment \cite{holzhutter2006thegeneralised}.
Since execution of flux minimisation in COBRApy requires a model in which reversible reactions are split into two irreversible reactions, a gsmodutils design was created in which all reversible reactions are split in a programmatic manner.
In the case of MetaCLAU, the flux minimal solution includes both ethanol and acetate production, which represents good qualitative agreement with experimental data (Norman et al., unpublished).
Since any changes to this predicted phenotype must be investigated, the flux minimisation analysis has been formulated as a gsmuodutils test.

\textbf{Product scans:}
Of interest to this project were changes in the product spectrum of \textit{C. autoethanogenum} under conditions where the organism can and cannot produce molecular hydrogen (with carbon monoxide as sole carbon and energy source).
The hypothesis tested in (Norman et al., unpublished), was that in the case where hydrogen production in infeasible, alternative electron sinks like lactate and 2,3BD would be produced.
As in the previous case, the model-predicted behaviour showing both lactate and 2,3BD was deemed an important result, which model curators should be notified of if lost during model development.
Thus the analysis was built into a gsmodutils test.

@@

% $\Delta \textit{nfn}$ knockout mutant: NOT SURE ABOUT THIS. CHRIS NOW THINKS IT'S ESSENTIAL :(
These test cases are all fully reproducible and available as an example within the gsmodutils software repository and we encourage the reader to explore these examples to understand how to utilise gsmodutils within their own project.

\section{Related work}
\textbf{Memote:} \cite{lieven2017memote} is tool set that has very similar ambitions to gsmodutils.
It features a fully specified set of tests, including custom test cases and has strong version control integration with git.
A core difference between these projects is that gsmodutils has a stronger focus on reducing the redundancy in model storage through the use of \textit{design deltas}, as described above.
Similarly, a core goal of gsmodutils is to allow easy import and export outside of the framework for compatibility with other modelling suites.
It should be noted that, as memote is written in python and utilises COBRApy, it is fully compatible with gsmodutils.
\\
\textbf{Cameo:} \cite{cardoso2017cameo} is a set of strain design utilities, such as genetic algorithms for finding optimal gene knock-out
strategies, as well as other flux balance analysis utilities. Written in python and part of the COBRApy suite of applications,
gsmodutils integrates easily with cameo.
Indeed, making a set of design changes can easily be integrated into gsmodutils through the python API.
\\
\textbf{Model repositories:} models are frequently shared, at the time of publication through services such as BiGG \cite{king2015bigg} and Biomodels \cite{chelliah2013biomodels}. 
Whilst these repositories encourage the reuse of models and the reproducibility of \textit{in silico} predictions they are not designed to improved collaboration.
The software presented here is designed with the notion that genome scale models are never finished, \textit{per se}, but under continuous development.
The cornerstone of this is the use of test cases, which formalise modelling validation criteria.

@@

As with most software development projects, gsmodutils will see expanded features.
Initially this will include tighter integration with version control systems such as git and mercurial.
Furthermore, the objective of the project is to cultivate collaboration by simplifying the process of distributing large models to different users.
As a consequence, methods for sharing reusable docker containers will be investigated.

>>>>

\title{Gsmodutils: A python based framework for test-driven genome scale metabolic model development}

@@

\begin{abstract}
\textbf{Motivation:} Genome scale metabolic models (GSMMs) are increasingly important for systems biology and metabolic engineering research as they are capable of simulating complex steady-state behaviour.
Constraints based models of this form can include thousands of reactions and metabolites, with many crucial pathways that only become activated in specific simulation settings.
However, despite their widespread use, power and the availability of tools to aid with the construction and analysis of large scale models, little methodology is suggested for the continued management of curated large scale models.
For example, when genome annotations are updated or new understanding regarding behaviour of is discovered models may need to be updated.
This is quickly becoming an issue for industrial systems and synthetic biotechnology applications, which require good quality reusable models integral to the design, build and test cycle.
\\
\textbf{Results:} 
As part of an ongoing effort to improve genome scale metabolic analysis, we have developed a test-driven development methodology for the continuous integration of validation data from different sources.
Contributing to the open source technology based around COBRApy, we have developed the \textit{gsmodutils} modelling framework placing an emphasis on test-driven design of models through defined test cases.
Crucially, different substrates are configurable allowing users to examine how different designs or curation impact a wide range of system behaviours, minimising error between model versions.\\
\textbf{Availability:} The software framework described within this paper is open source and freely available from \url{http://github.com/SBRCNottingham/gsmodutils} \\
\textbf{Contact:}{james.gilbert@nottingham.ac.uk}
\end{abstract}

\section{Introduction}
Stoichiometric constraints based modelling for biological systems has been a mainstay of systems biology for several decades \cite{fell1986fat, varma1994stoichiometric}.
Given its flexibility, low barrier to entry and requirement only on minimal knowledge regarding the stoichiometry of metabolic networks this structural approach has become an extremely popular method for modelling steady-state behaviour of large, biochemical networks \cite{kauffman2003advances}.
Such large scale reconstructions are often referred to as genome scale metabolic models (GSMMs), as the processes is significantly aided through the advent of relatively inexpensive genome sequencing \cite{o2015using, land2015insights}.
Indeed, owing to their ability to model complex aspects of metabolism, GSMMs have been widely adopted as a standard to elucidate and optimise of industrial biotechnology processes \cite{kim2017current}.

@@

As a consequence, a significant amount of work has gone into the management of genome scale models.
The BiGG models database \cite{king2015bigg}, for example, exists to provide a standardised repository of validated models that can be shared and reused.
However, little focus is placed upon the collaborative design aspect of such models with few mechanisms existing to capture the differences between two model versions, \textit{model deltas}.
Similarly, the MetaNetX \cite{moretti2016metanetx} system exists to provide a standardised namespace and toolchain for GSM analysis.
However, such tools often make it difficult to understand the design decisions made by the initial model authors.

Furthermore, as with many areas of bioinformatic study the number available computational tools has become vast.
This covers huge variety of software platforms including the COBRA toolbox for Matlab \cite{schellenberger2011quantitative}, ScrumPy and COBRApy in python \cite{poolman2006scrumpy, ebrahim2013cobrapy} with additional tools and libraries such as cameo \cite{cardoso2017cameo}, OptFlux in Java \cite{rocha2010optflux} and SurreyFBA \cite{gevorgyan2010surreyfba}.
Whilst most of these tools are Open Source and follow standards, such as SBML \cite{finney2003systems}, it is often challenging to replicate the initial modelling efforts conducted by authors of papers.
Consequently, we feel that software tools are urgently needed to address this issue.

Similarly, The archetype design, build and test cycle of synthetic biology heavily relies on the usage of bioinformatics software and modelling to improve the production of natural products \cite{carbonell2016bioinformatics}.
In order to speed up the use of bioinformatics tools to produce high value platform chemicals, genome scale models are often used to discover methods for process optimisation \textit{in silico}.

For example, many tools such as RetroPath \cite{carbonell2013retropath}, XTMS \cite{carbonell2014xtms} and GEM-Path \cite{campodonico2014generation} suggest thousands of potential heterologous pathways.
Many of these tools significantly increase the value of genome scale models, for example by coupling commodity production to an organism's growth \cite{feist2010model}.
These tools all suggest major changes to wild-type strains must be tracked and compared to allow models to remain relevant.
In effect, mechanisms are required to relate to modified test and production strains.

Similarly, many conventional applications of genome scale models in systems biology have often suffered from unnecessary replication of work due to a lack of adherence to standards \cite{monk2014}.
For example, there are now many independently developed models of \textit{Clostridium acetobutylicum} \cite{lee2008genome, senger2008genome, mcanulty2012genome, dash2014capturing, yoo2015quantitative}, an organism used in the production of solvents for around a century \cite{moon2016one, 1919charles}.
These models all exist to solve similar biological problems some being updates to the inital base models. 
However, there has been disagreement over fundamental biochemical properties of this anaerobic organism, notably with the focus on redox balancing \cite{dash2014capturing}.
Such models also include updates based on improved genome annotations and the inclusion of fluxomic, transcriptomic and metabolomic characterisations \cite{yoo2015quantitative}.

%Many of the results presented in original sources are difficult to compare due to a lack of standardisation of reference identifiers for reaction names.

Unfortunatley, many of the results reported in \cite{lee2008genome, senger2008genome, mcanulty2012genome, dash2014capturing, yoo2015quantitative} are difficult to compare or reproduce as the result of a number of issues.
Often, the models lack a standardisation of identifiers for reaction names, which makes direct comparisson of model structure as well as differences between reactions a challenge.
Where models are shared, it is often in non-standard spreadsheet formats, rather than SBML models.
Indeed, even in the case of valid SBML models being made available at the time of publication few details are given as to how to run such models for conditions discussed in original articles.

In this paper, we present a software framework geared toward \textit{test-driven} genome scale model development, a concept that is taken directly from good software development practices \cite{martin2002agile}.
By this we mean the notion that, as a model is curated to represent biological phenomena, much of the validation can be turned into specific test cases that can be repeated between model versions.
We provide an example test case for \textit{Clostridium autoethanogenum}, an organism that has had considerable focus in terms of genome scale models and how a working methodology using the software presented here can reduce repetition of work and improve the reproducibility of results.
This article aims to summarise the main objectives of the \textit{gsmodutils} software and we refer the reader to the software user guide for a more detailed exploration features.

@@

In contrast, an \textit{agile} methodology for the development of models places the focus on adapting work to new requirements \cite{martin2002agile}.
Such an approach best fits genome scale models because they are rarely created to investigate individual processes and, instead, capture the complexity of large systems.
Genome scale models are intrinsically related to available genome annotations.
Such annotations rely heavily on automated matches to related species, with the characterisations of individual genes often being left to a few of critical interest \cite{seemann2014prokka}.
This modelling formation costs an in depth understanding of dynamic behaviour. 
However, capturing steady-state phenomena still provides a good understanding of system properties \cite{o2015using}.

@@

Therefore, models undergo significant manual annotation and curation; a process which has a high chance of error.
In this work we advocate a test-driven approach to model development highlighted in Figure \ref{fig:test_driven}.
Here, the model is changed to achieve core research goals that are dynamic in response to the changes of a project.
In order to meet this objective, validation criteria for a model such as growth conditions or the impact of gene knock outs, should be formally set.
When a model is changed, all such validation criteria should be retested to ensure that models do not regress to previous states.

@@

We feel that many of the current software tools for genome scale models do an excellent job of facilitating answers to crucial research and design questions.
However, there is a major gap in terms of the reliability and re-usability of models due to a lack of standardisation and software tools to aid such processes.
The following sections provide an overview of the \textit{gsmodutils} software framework.
\textit{gsmodutils} aims to provide a basis for test-driven, version controlled agile model development.
All software is open source and is designed to be interoperable with platforms widely used in the domain of constraints based modelling.

@@

\subsection{Outline and features}
Test-driven development is driven by the idea of clearly defined test cases written before significant changes are made to any underlying architecture.
In the case of genome scale models, errors occur as a product of human curation to better represent newly discovered aspects of metabolism.

By automatically integrating COBRApy \cite{ebrahim2013cobrapy} users can easily write convenient test cases following examples given in the user guide.
A standard test case, ensuring that a given model grows on media is given in Figure \ref{fig:test_case}.
When a new model repository is created with the \textit{gsmodutils} tool, a number of pre-written test cases are automatically added to a file.
However, we stress that the vast majority of individual use cases for a model must be specific to a given biological problem.

@@

\lstset{language=python}   
\begin{lstlisting}
def test_func(model, project, log):
    status = model.solver.optimize()
    log.assertion(status == 'optimal',
                  ``Works'', # Success message
		  ``Does not work'', # Failure message
                  ``Test description'')
\end{lstlisting}
\caption{An example \textit{gsmodutils} test case written in python}
\label{fig:test_case} 
\end{figure}

\subsection{Strain designs}
A core aspect behind the implementation of \textit{gsmodutils} is the concept of a \textit{design}, this encompasses a simple set of changes to a ``wild-type'' model that are required for analysis.
For example a mutant strain that has undergone several gene deletions is not sufficiently different from the original wild-type to merit having its own genome scale model.
However, it is often the case that such deletions are of scientific or industrial interest and, as such, the strain will be used in future work.
Consequently, such designs are hereditary in nature.
By taking the \textit{model delta} between the constraints applied to an initial model and subsequent modifications, \textit{gsmodutils} allows users to easily reuse and export models with this design.
As designs inherit from a base model, future curation to a wild type model will automatically be included in the designs.
Similarly, designs are self contained and will not interfere with one another allowing a form of project management.
Figure \ref{fig:strain_hered} shows how this could work in a practical situation.

Designs of this nature can also be programmatic, allowing the implementation of features such as non-standard constraints that can be dynamically loaded.
An example of this is shown in Figure \ref{fig:programmatic_design}.
This example converts an existing model to one based on a Mixed Integer Linear Program (MILP), as the reaction names do not need to be specified, should reactions be added or removed the process will still apply.
Alternative examples could include reductions of models through methods such as elementary flux modes or minimum cut sets, which can change dramatically with only small changes to stoichiometry.
Furthermore, integrity of all strain designs is automatically tested as part of the default \textit{gsmodutils} testing framework.

\begin{figure}[ht]
\lstset{language=python}   
\begin{lstlisting}
def gsmdesign_example_milp(model, project):
    for reaction in model.reactions:
	reaction.variable_kind = "integer"
    return model
\end{lstlisting}
\caption{An example \textit{gsmodutils} programmatic design written in python. 
This design converts reactions to integer type, allowing an MILP formation.}
\label{fig:programmatic_design} 
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{inheritence2.eps}
  \caption{Examples of \textit{gsmodutils} design inheritance.
  Each design stores the delta between the wild-type base model, any parents and the changes to constraints the design contains.
  In the example presented above a heterologous production pathway is combined with a reusable set of knock-outs.
  In practice, ideally, these designs should relate to real constructs and strains evaluated by experiment.
  }
    \label{fig:strain_hered}
\end{figure}

\subsection{Development workflow}
In this section we propose a method for the development of genome scale models that integrates \textit{gsmodutils} with version control systems.
Figure \ref{fig:test_case} highlights the notion of test cases, taken from test-driven development.
The outlined workflow is that the user writes a formal test case for some modelling goal, perhaps driven by captured experimental data, that fits a specific form of validation criteria.
We note that, in principle, this test case should be written before changes to a model.

\section{Case study usage Clostridium autoethanogenum}
\textit{Clostridium autoethanogenum} is a bacterial species used for the production of commodity chemicals at industrial scale \cite{abrini1994clostridium,norman2018progress,LT}. A new GSMM of \textit{C. autoethanogenum}, `MetaCLAU', has been analysed to improve this bioprocess (Norman et al., submitted for peer review).
In this section we describe how \textit{gsmodutils} has been utilised to ensure that future versions of MetaCLAU will remain functionally relevant from the perspective of industrial biotechnology.

@@

\textbf{Minimisation of flux based tests:}
One conventional approach in FBA is to set an optimisation criterion of minimising flux across enzymatic reactions with a fixed biomass constraint \cite{holzhutter2006thegeneralised}. 
The solution to this FBA problem represents minimal protein investment \cite{holzhutter2006thegeneralised}.
Since execution of flux minimisation in COBRApy requires a model in which reversible reactions are split into two irreversible reactions, a \textit{gsmodutils} design was created in which all reversible reactions are split in a programmatic manner.
In the case of MetaCLAU, the flux minimal solution includes both ethanol and acetate production, which represents good qualitative agreement with experimental data (Norman et al., unpublished).
Since any changes to this predicted phenotype must be investigated, the flux minimisation analysis has been formulated as a gsmuodutils test.

\textbf{Product scans:}
Of interest to this project were changes in the product spectrum of \textit{C. autoethanogenum} under conditions where the organism can and cannot produce molecular hydrogen (with carbon monoxide as sole carbon and energy source).
The hypothesis tested in (Norman et al., unpublished), was that in the case where hydrogen production in infeasible, alternative electron sinks like lactate and 2,3BD would be produced.
As in the previous case, the model-predicted behaviour showing both lactate and 2,3BD was deemed an important result, which model curators should be notified of if lost during model development.
Thus the analysis was built into a \textit{gsmodutils} test.

@@

% $\Delta \textit{nfn}$ knockout mutant: NOT SURE ABOUT THIS. CHRIS NOW THINKS IT'S ESSENTIAL :(
These test cases are all fully reproducible and available as an example within the \textit{gsmodutils} software repository and we encourage the reader to explore these examples to understand how to utilise \textit{gsmodutils} within their own project.

\section{Related work}
\textbf{Memote:} \cite{lieven2017memote} is tool set that has very similar ambitions to gsmodutils.
It features a fully specified set of tests, including custom test cases and has strong version control integration with git.
A core difference between these projects is that \textit{gsmodutils} has a stronger focus on reducing the redundancy in model storage through the use of \textit{design deltas}, as described above.
Similarly, a core goal of \textit{gsmodutils} is to allow easy import and export outside of the framework for compatibility with other modelling suites.
It should be noted that, as memote is written in python and utilises COBRApy, it is fully compatible with gsmodutils.
\\
\textbf{Cameo:} \cite{cardoso2017cameo} is a set of strain design utilities, such as genetic algorithms for finding optimal gene knock-out
strategies, as well as other flux balance analysis utilities. Written in python and part of the COBRApy suite of applications,
gsmodutils integrates easily with cameo.
Indeed, making a set of design changes can easily be integrated into \textit{gsmodutils} through the python API.
\\
\textbf{Model repositories:} models are frequently shared, at the time of publication through services such as BiGG \cite{king2015bigg} and Biomodels \cite{chelliah2013biomodels}. 
Whilst these repositories encourage the reuse of models and the reproducibility of \textit{in silico} predictions they are not designed to improved collaboration.
The software presented here is designed with the notion that genome scale models are never finished, \textit{per se}, but under continuous development.
The cornerstone of this is the use of test cases, which formalise modelling validation criteria.

@@

As with most software development projects, \textit{gsmodutils} will see expanded features.
Initially this will include tighter integration with version control systems such as git and mercurial.
Furthermore, the objective of the project is to cultivate collaboration by simplifying the process of distributing large models to different users.
As a consequence, methods for sharing reusable docker containers will be investigated.
